---
title: "MDS Assignment 12"
author: "M Meyer (22675760)"
date: "2024-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(StatMatch)
library(MASS)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(cluster)
library(combinat)
library(maps)

melanoma <- data.frame(h = c(22,16,19,11), t = c(2,54,33,17), e = c(10,115,73,28), row.names = c("H", "S", "N", "I"))

```

## Assignment 12.1

### (a) Show that the above ð‘¼, ð‘½ and $D_r$ provide the GSVD of ð‘¨. 

### (b) Write an R function to calculate the GSVD of a rectangular matrix.

The code below uses the function for the symmetric square root matrix that was formed for Assignment 2.

```{r}
symSqrt <- function(A) {
  # Perform eigenvalue decomposition
  eigDecomp <- eigen(A)
  
  # Get the eigenvectors and eigenvalues
  eigVecs <- eigDecomp$vectors
  eigVals <- eigDecomp$values
  
  # Calculate the square root of the eigenvalues
  sqrtEigVals <- sqrt(eigVals)
  
  # Form the symmetric square root matrix
  sqrtMat <- eigVecs %*% diag(sqrtEigVals) %*% t(eigVecs)
  
  return(sqrtMat)
}

myGSVD <- function(dat=melanoma)
{
  A <- as.matrix(A)
  Dr <- diag(apply(A, 1, sum))
  Dc <- diag(apply(A, 2, sum))
  Omega <- solve(Dr)
  Phi <- solve(Dc)
  Osqrt <- symSqrt(Omega)
  Phisqrt <- symSqrt(Phi)
  
  N <- (svd(Osqrt %*% A %*% Phisqrt))$u
  M <- (svd(Osqrt %*% A %*% Phisqrt))$v
  D <- diag(round((svd(Osqrt %*% A %*% Phisqrt))$d, 3))
  
  U <- solve(Osqrt) %*% N
  V <- solve(Phisqrt) %*% M
  
  return(list(round(U,4), D, round(V, 4)))
}

melanomaResults <- myGSVD(melanoma)

# > round(t(U)%*%Omega%*%U, 2)
#      [,1] [,2] [,3]
# [1,]    1    0    0
# [2,]    0    1    0
# [3,]    0    0    1
# > round(t(V)%*%Phi%*%V, 2)
#      [,1] [,2] [,3]
# [1,]    1    0    0
# [2,]    0    1    0
# [3,]    0    0    1
```

## Assignment 12.2

### Make use of the R function written in 12.1 for calculating the GSVD to obtain the correspondence analysis of C&C Table 9.1 and compute the chi-square distances between the rows as well as between the columns.

## Assignment 12.3

### Carry out a correspondence analysis on C&C Table 9.1. Your R function must provide for displaying the row space and column space on separate diagrams as well as on the same diagram; the total inertia must be calculated as well as the relative contributions of the different dimensions to the total inertia.

```{r}

```
