---
title: "MDS Assignment 8"
author: "M Meyer (22675760)"
date: "2024-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(StatMatch)
library(MASS)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(cluster)
library(combinat)

# import bioenv data:
# setwd("~/Desktop/Multi-Dim Scaling/7")
cereal <- read.csv("Kellog.csv", header=TRUE)
cerealNum <- cereal[,-1]
```



## Assignment 7.1

### (a) 	Note the steps for applying Procrustes analysis in practice, C&C 5.2.1

Done in own time

### (b) 	Write an R-function to implement the four steps of an orthogonal Procrustes-analysis in practice.

```{r}

myProcrustes <- function(X,Y)
{
  
  q <- ncol(X)
  p <- ncol(Y)
  n <- nrow(X)
  
  # 1- padding to ensure the same nr of columns
  if (q < p)
  {
    while(ncol(X) < ncol(Y))
    {
      X <- cbind(X,0)
    }
  } else if (q > p) {
    while(ncol(Y) < ncol(X))
    {
      Y <- cbind(Y,0)
    }
  }
  
  # 2- column centre both of them
  Xc <- apply(X, 2, function(x) x - mean(x))
  Yc <- apply(Y, 2, function(y) y - mean(y))
  
  # 3- find svd of Y'X
  YX <- svd(t(Yc)%*%Xc)
    
  L <- diag(YX$d)
  U <- YX$u
  V <- YX$v
  
  # 4- A=VY'
  A <- V%*%t(U)
  
  # 5- Calculate rho
  XAY <- Xc%*%A%*%t(Yc)
  XX <- t(Xc)%*%Xc
  rho <- sum(diag(XAY))/sum(diag(XX))
  
  fittedX <- rho*Xc%*%A
  
  stat_num <- (sum(L))^2
  stat_denom <- sum(diag(t(Xc)%*%Xc))*sum(diag(t(Yc)%*%Yc))

  procrustesStat <- 1-stat_num/stat_denom
  
  return(list(centred_Y = Yc, centred_X = Xc, fitted_X = fittedX, procrustesStat = procrustesStat))
}

```

### (c) Obtain a classical scaling of the Breakfast cereal data as well as a non-metric scaling of the same data. Use the R-function in (b) to fit the non-metric MDS configuration of the Breakfast cereal data to the classical MDS display.

```{r, fig.asp=1}
# scaleCer <- scale(cerealNum)
# Scale each variable to have range (0,1)
# scaleCer <- apply(cerealNum, 2, function(x) (x - min(x)) / (max(x) - min(x)))
# cerDist <- as.matrix(dist(scaleCer))

cerealNum$Shelf <- as.factor(cerealNum$Shelf)
cerealNum$VitMins <- as.factor(cerealNum$VitMins)
cerDist <- daisy(cerealNum, metric = "euclidean", stand=TRUE)


Y <- cmdscale(cerDist) #target - classic
X <- (isoMDS(cerDist))$points #testee - non-metric

result <- myProcrustes(X=X, Y=Y)
result$fitted_X
result$procrustesStat

par(mfrow = c(2,2))

plot(result$centred_Y, type = "n", main = "Classical scaling (Target)", xlab = "", ylab = "")
text(result$centred_Y, labels = cereal$X, cex=0.7)

plot(result$centred_Y, type = "n", main = "Classical vs fitted Non-metric", xlab = "", ylab = "")
text(result$centred_Y, labels = cereal$X, cex=0.7, col = "darkred")
text(result$fitted_X, labels = cereal$X, cex=0.7, col = "darkturquoise")
legend("bottomleft", legend = c("Classical", "NonMetric"), col = c("darkred","darkturquoise"), 
       bty = "n",cex = 0.5, lty = 1)

plot(result$centred_X, type = "n", main = "Original non-metric scaling (Testee)", xlab = "", ylab = "")
text(result$centred_X, labels = cereal$X, cex=0.7)

plot(result$fitted_X, type = "n", main = "Fitted non-metric scaling (Testee)", xlab = "", ylab = "")
text(result$fitted_X, labels = cereal$X, cex=0.7, lty = 1)

par(mfrow = c(1,1))
```

