---
title: "Assignment 4 - MDS"
author: "M Meyer (22675760)"
date: "2024-03-8"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# import skulls data:
pine <- read.csv("Pinedata.csv", header=TRUE)
pine1 <- pine[,-(1:2)]

library(MASS)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(optiscale)
library(stats)

pineCols <- rep("orange", 36)
pineCols[pine$species=="P. kes"] <- "turquoise2"
pineCols[pine$species=="P. max"] <- "green2"
pineCols[pine$species=="P. pat"] <- "magenta"
pineCols[pine$species=="P. tae"] <- "yellow2"
```

# Assignment 4.1

#### Write R code to construct an MDS display for an AoD analysis. Provision must be made for plotting the group means as well as all the samples.

Use classical scaling to obtain $Λ_p$ and $\overline{\underline X}$.


```{r, collectcode=TRUE}
# Classical scaling function:

classScale <- function(D)
{
  ddistMat <- D
  n <- nrow(D)
  centeringMat <- diag(n)-1/n
  B <- centeringMat %*% ddistMat %*% centeringMat
  
  allEigen <- eigen(B)$values
  eigenVals <- allEigen[allEigen>0]
  q <- length(eigenVals)
  
  V <- as.matrix(eigen(B)$vectors[,1:q])
  
  xStar <- V%*%diag(sqrt(eigenVals))
  return(list(xStar=xStar, Lp=eigenVals))
}

# Analysis of distance
aod <- function(data=pine1, p=2, groups=pine$species)
{
  n <- nrow(data)
  K <- length(unique(groups))
  # scale the data:
  Z <- scale(data)
  # distance matrix:
  D <- as.matrix(dist(Z, diag=TRUE))
  # Create G:
  groups.int <- as.integer(as.factor(groups))

  G <- matrix(0, nrow=n, ncol = K)
  for (i in 1:n)
  {
    G[i,groups.int[i]] <- 1
    }
  GG.inv <- diag(1/apply(G,2,sum))
  # Calculate Phi
  Phi <- -0.5*D^2
  # Calculate Theta
  Theta <- GG.inv%*%t(G)%*%Phi%*%G%*%GG.inv
  # Only need theta, because CS the same as for Delta when double-centering
  vec1 <- rep(1,K)
  # PERFORM CLASSICAL SCALING:
  Xstar <- ((classScale(Theta))$xStar)[,1:2] # 2 dimensions
  Lp <- diag(((classScale(Theta))$Lp)[1:2])
  
  # Supplementary point coordinates:
  d0 <- Xstar[,1]^2 + Xstar[,2]^2 # squared distance to origin
  
  # Unsure of how to calculate d_i...
  # First perform classical scaling of original dataset?
  Zi <- cmdscale(dist(scale(data)))
  
  di <- matrix(0, nrow=nrow(data), ncol = length(d0))
  for (i in 1:nrow(data))
  {
    for (j in 1:length(d0))
    {
      di[i,j] <- (Zi[i,1]-Xstar[j,1])^2 + (Zi[i,2]-Xstar[j,2])^2 # squared distance to group coordinate
    }
  }
  Xi <- matrix(0, nrow=nrow(data), ncol = 2)
  s <- apply(G,2,sum)/nrow(data)
  for (l in 1:nrow(data))
  {
    Xi[l,] <- -0.5*solve(Lp) %*% t(Xstar) %*% (diag(1, nrow=length(d0))-vec1%*%t(s)) %*% (di[l,]-d0)
  }
  plot(Xi, col = pineCols, cex = 0.7, main = "MDS display of AoD Analysis",
       xlab = "X", ylab = "Y")
  points(Xstar, pch=16, col = c("orange", "turquoise", "green2", "magenta", "yellow2"), cex = 1.5)
  legend("topright", legend = unique(groups), col = c("orange", "turquoise", "green2", "magenta", "yellow2"), pch = 21 , cex = 0.8)
}

```

## Assignment 4.2 

#### (a)	Perform an AOD of the given data with Euclidean distances between the group means.

Please note that with the changing use of notation in the notes, I suspect there might be an error in the function.

I was also unsure as how to calculate $\underline d_i$.

```{r}
aod()
```

#### (b)	Write the necessary R code to perform a permutation test on AoD data. Now, use your program to use a permutation test to test if the five pine species differ significantly from one another. Calculate the ASL (achieved significance level) and display graphically.

```{r}
permTestF <- function(data=pine1, p=2, groups=pine$species)
{
  n <- nrow(data)
  # scale the data:
  Z <- scale(data)
  # distance matrix:
  D <- as.matrix(dist(Z, diag=TRUE))
  
  K <- length(unique(groups))
  # Create G:
  groups.int <- as.integer(as.factor(groups))

  G <- matrix(0, nrow=n, ncol = K)
  for (i in 1:n)
  {
    G[i,groups.int[i]] <- 1
    }
  G
  GG.inv <- diag(1/apply(G,2,sum))
  # Calculate Phi
  Phi <- -0.5*D^2
  # Calculate Theta
  Theta <- GG.inv%*%t(G)%*%Phi%*%G%*%GG.inv
  Theta.diag <- diag(diag(Theta))
  # Delta:
  vec1 <- rep(1,K)
  Delta <- Theta-0.5*(Theta.diag%*%vec1%*%t(vec1) + vec1%*%t(vec1)%*%Theta.diag)
  
  n.vec <- apply(G,2,sum)
  
  B <- t(n.vec) %*% Delta %*% n.vec
  
  W <- numeric(K)
  
  for (j in 1:K)
  {
    W[j] <- (n/n.vec[j]) * G[,j] %*% Phi %*% G[,j]
  }
  W <- sum(W)
  
  Fstat <- (B/(K-1))/(W/(n-K))
  return(as.double(Fstat))
}

# Create random permutations of the data's group allocations
N <- 2000
Fstats <- numeric(N)

set.seed(12345)
for (i in 1:N)
{
  randomGroup <- sample(as.integer(as.factor(pine$species)), nrow(pine), replace=FALSE)
  randomSpecies <- unique(pine$species)[randomGroup]
  Fstats[i] <- permTestF(groups=randomSpecies)
}

F0 <- permTestF()

ggplot(data = data.frame(x = Fstats), aes(x = x)) +
  geom_histogram(binwidth = 0.25, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_vline(xintercept = F0, color = "firebrick", size=0.75,linetype = "dotted", alpha=0.7) + 
  annotate("text", x = F0 + 0.1, y = 150, label = "F0", color = "firebrick", size = 4) +
  labs(title = "Histogram of F statistics",
       x = "F Statistic",
       y = "Frequency")

cat("Achieved significance level is ", round(sum(Fstats>F0)/N, 5))
```












