---
title: "MDS Assignment 11"
author: "M Meyer (22675760)"
date: "2024-04-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(StatMatch)
library(MASS)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(cluster)
library(combinat)
library(maps)

# setwd("~/Desktop/Multi-Dim Scaling/9")
Nations <- read.csv("Nations.csv", header=TRUE, row.names = 1)

```

## Assignment 1.1

(a) Implement the algorithm given in C&C section 5.2.2 in an R-function.

### (a) Write an R function to implement the ALS-unfolding algorithm of Greenacre and Browne.

The following algorithm performs 2-dimensional unfolding.
Delta is an $\Delta: n\times m$ matrix.

```{r}
alsUnfold <- function(delta)
{
  delta2 <- delta^2

  # 1. initialize
  td <- -0.5 * (delta^2) # delta tilde
  n <- nrow(delta)
  m <- ncol(delta)
  
  In <- diag(1,n)
  Im <- diag(1,m)
  m1 <- matrix(rep(1/m, m^2), ncol = m)
  n1 <- matrix(rep(1/n, n^2), ncol = n)
  
  C <- (In-n1)%*%td%*%(Im-m1)
  svdC <- svd(C, nu = n, nv = m)
  Ltemp <- matrix(0, ncol = m, nrow= n)
  diag(Ltemp) <- svdC$d
  svdC$u%*%Ltemp%*%t(svdC$v)
  
  l1 <- sqrt(svdC$d[1])
  l2 <- sqrt(svdC$d[2])
  temp0 <- matrix(0, nrow=m, ncol = 2)
  diag(temp0) <- c(l1, l2)
  Y0 <- svdC$v%*%temp0
  Y <- Y0
  
  stresses <- numeric(1)
  stress1 <- 99999999999
  j = 1
  
  # START PERM HERE!!
  
  while (j < 1000)
  {
    stress0 <- stress1
    
    L <- t(Y)%*%Y
    L1 <-  L[1,1]
    L2 <- L[2,2]
    
    # 2 - Update X
    
    # Y0: m x 2 = 21 x 2
    # Y0: 2 x m = 2  x 21
    # delta r = 21 x 1
    # h = 21 x 1 (diag as in a vector)
    w <- matrix(0, ncol = n, nrow = 2)
    c <- numeric(n)
    psi <- matrix(0, ncol = n, nrow = 6)
    psi[6,] <- 1
    # 2 x 21 . (21 x 1)
    h = diag(Y%*%t(Y))
    for (r in 1:n)
    {
      w[,r] <- t(Y)%*%(delta2[r,]-h)
      w1 <- w[1,r]
      w2 <- w[2,r]
      c[r] <- sum(delta2[r,]-h)
      psi[5,r] <- -(4*(L1+L2)+c[r])
      psi[4,r] <- 4*( (L2^2 + 4*L1*L2 + L1^2) + c[r]*(L1 + L2) )     
      psi[3,r] <- m* (w1^2 + w2^2) + -4*(c[r]*(L1^2 + 4*L1*L2 + L2^2) + 4 *(L1*L2^2 + L2*L1^2))   
      #psi[3,r] <- m*( (w1^2 + w2^2) + -4*(c[r]*(L1^2 + L2^2) + 4 *(L1*L2^2 + L2*L1^2-L1*L2)))   
      psi[2,r] <- 4 *(4*((L1^2)*(L2^2) + c[r]*(L1*L2^2 + L2*L1^2)) - m* (L2*w1^2 + L1*w2^2)) 
      psi[1,r] <- 4 * m * ((w1^2)*L2^2 + (w2^2)*L1^2) - c[r]*16*(L1^2)*(L2^2)
    }
    psi
    psiroots = apply(psi,2,polyroot )
    minroots = apply(Re(psiroots),2,min)
    minroots
    
    XjCol1 <- w[1,]/(minroots-2*L1)
    XjCol2 <- w[2,]/(minroots-2*L2)
    
    X <- cbind(XjCol1,XjCol2)
    X <- scale(X,center = TRUE, scale = FALSE)
    
    svdX <- svd(X)
    Ux <- svdX$u
    dx <- diag(svdX$d)
    
    X <- Ux%*%dx
    
    xL <- t(X)%*%X
    xL1 <-  xL[1,1]
    xL2 <- xL[2,2]
    
    # 3 - Update Y
    
    # X: 18 x 2
    # delta i = 18 x 1 (columns of delta)
    # h = 18 x 1 (diag as in a vector)
    xw <- matrix(0, ncol = m, nrow = 2)
    xc <- numeric(m)
    xpsi <- matrix(0, ncol = m, nrow = 6)
    xpsi[6,] <- 1
    # 2 x 21 . (21 x 1)
    xh = diag(X%*%t(X))                       #EDIT
    for (i in 1:m)                              # EDIT
    {
      xw[,i] <- t(X)%*%(delta2[,i]-xh)
      xw1 <- xw[1,i]
      xw2 <- xw[2,i]
      xc[i] <- sum(delta2[,i]-xh)
      xpsi[5,i] <- -(4*(xL1+xL2)+xc[i])
      xpsi[4,i] <- 4*( (xL2^2 + 4*xL1*xL2 + xL1^2) + xc[i]*(xL1 + xL2) )     
      xpsi[3,i] <- n* (xw1^2 + xw2^2) + -4*(xc[i]*(xL1^2 + 4*xL1*xL2 + xL2^2) + 4 *(xL1*xL2^2 + xL2*xL1^2))
      xpsi[2,i] <- 4 *(4*((xL1^2)*(xL2^2) + xc[i]*(xL1*xL2^2 + xL2*xL1^2)) - n* (xL2*xw1^2 + xL1*xw2^2)) 
      xpsi[1,i] <- 4 * n * ((xw1^2)*xL2^2 + (xw2^2)*xL1^2) - xc[i]*16*(xL1^2)*(xL2^2)
    }
    xpsi
    xpsiroots = apply(xpsi,2,polyroot )
    xminroots = apply(Re(xpsiroots),2,min)
    xminroots
    
    YjCol1 <- xw[1,]/(xminroots-2*xL1)
    YjCol2 <- xw[2,]/(xminroots-2*xL2)
    
    Y <- cbind(YjCol1,YjCol2)
    
    stressFun <- function(x,y) {
      z <- numeric(nrow(y))
      for (j in 1:(nrow(y))) {
        z[j] <- t(x-y[j,])%*%(x-y[j,])
      }
      return(z)
    }
    
    stressInner <- delta2 - t(apply(X, 1, stressFun, Y))
    stress <- sum( stressInner^2 )
    stress1 <- stress
    stresses <- append(stresses, stress1)
    
    # terminate if converge
    if (abs(stress0 - stress1) < 0.0001) {
      break
      }
    
    # else, perform PCA on Y and iterate
    
    j <- j+1
    
    Y <- scale(Y,center = TRUE, scale = FALSE)
    
    svdY <- svd(Y)
    Uy <- svdY$u
    dy <- diag(svdY$d)
    
    Y <- Uy%*%dy  
    
    # now just save the first X and Y to check with prof Lubbe
    if (j==2) {
      X_1 <- X
      Y_1 <- Y
    }
  }
  return(list(Y = Y, Y1 = Y_1, X1 = X_1, X=X, finalStress = tail(stresses,1), StressPath = stresses, iters = j))
}

```

### (b) Describe the Nations dataset of Wish et al. (1972).

This data set represents the ratings given by students to measure the similarity between pairs of nations. Every nation was given a score based on 18 variables with a scale of 1 to 9. 

### (c) Perform a multidimensional unfolding on the nations data set with the above R function.

The results below agree with that from Cox & Cox and the results are illustrated below, replicating Figure 8.4(i) and 8.4(ii).

```{r, fig.asp=1}
# delta is given
delta <- as.matrix(Nations)
unfoldResults <- alsUnfold(delta)

natnames <- colnames(delta)
varnames <- row.names(delta)

y <- unfoldResults$Y
x <- unfoldResults$X

# PRINT OUT THE STRESS SUMMARY

stressSummary <- data.frame(stress = c(unfoldResults$StressPath[2], median(unfoldResults$StressPath), tail(unfoldResults$StressPath,1)))
row.names(stressSummary) <- c("Initial", "Median iteration", "Final")

myTable = knitr::kable(stressSummary, col.names=c("Value"), caption="Summary of stress over iterations",escape=FALSE, digits=4)
myTable %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# PRINT OUT THE FINAL Y 

finalY <- data.frame(Y = y)
row.names(finalY) <- natnames

myTableY = knitr::kable(finalY, col.names=c("x-coord", "y-coord"), caption="Final Y configuration (nations)",escape=FALSE, digits=4)
myTableY %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# PRINT OUT THE FINAL Y 

finalX <- data.frame(X = x)
row.names(finalX) <- varnames

myTableX = knitr::kable(finalX, col.names=c("x-coord", "y-coord"), caption="Final X configuration (variables)",escape=FALSE, digits=4)
myTableX %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

# CREATE PLOTS OF THE RESULTS

plot(y, type = "n", main = "(i) - Unfolding analysis of the rating of nations", xlab = "", ylab = "",
     xlim = c(min(min(y[,1]), min(x[,1])), max(max(y[,1]), max(x[,1]))),
     ylim = c(min(min(y[,2]), min(x[,2])), max(max(y[,2]), max(x[,2]))))
text(y, labels = natnames, cex=0.7, col = "darkred")
text(x, labels = varnames, cex=0.7, col = "darkturquoise")
legend("topright", legend = c("Nations", "Variables"), col = c("darkred","darkturquoise"), 
       bty = "n",cex = 0.75, lty = 1)

plot(y, type = "n", main = "(ii) - Magnifying the region occupied by the countries in (i)", xlab = "", ylab = "")
text(y, labels = natnames, cex=0.7, col = "darkred")

```

