---
title: "MDS Assignment 5"
author: "M Meyer (22675760)"
date: "2024-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# import skulls data:
kellog <- read.csv("Kellog.csv", header=TRUE)
kellog1 <- kellog[,-1]

library(MASS)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(cluster)

# pineCols <- rep("orange", 36)
# pineCols[pine$species=="P. kes"] <- "turquoise2"
# pineCols[pine$species=="P. max"] <- "green2"
# pineCols[pine$species=="P. pat"] <- "magenta"
# pineCols[pine$species=="P. tae"] <- "yellow2"
```

# Assignment 5.1

### (c) Consider the Kellogg breakfast cereals dataset as given by C&C: 

#### Obtain a distance matrix based upon Gower's general dissimilarity coefficient. (Hint: Study the help file of the R function `daisy()` in package `cluster`). 

```{r, collectcode=TRUE}
# first scale the kellog dataset before doing any further calculations:
kellog1$Shelf <- as.factor(kellog1$Shelf)
kellog1$VitMins <- as.factor(kellog1$VitMins)
kdist <- daisy(kellog1, metric = "gower")
kdist1 <- daisy(kellog1, metric = "gower", stand=TRUE)
# round(kdist,3)

```

#### Perform a non-metric MDS using `isoMDS()`. Compare results with those of C&C.

From the results below, we can see that the stress agrees with that of the textbook, being $15\%$. Even though the plot looks different than Figure 3.2 from the textbook, the configurations agree and in both of them the All Bran cereals are a bit separate in the one corner (bottom left below) and the corn products are close to each other (Cornflakes, Special K, Frosted Flakes and Corn Pops).

```{r, collectcode=TRUE}
kellogMDS <- isoMDS(kdist)

plot(kellogMDS$points, type = "n", ylab = "", xlab = "",
     main = "Non-metric scaling of Kellogg breakfast cereals")
text(kellogMDS$points, labels = kellog$X, cex=0.8)
```

### (d)	Use the R function Shepard() in package MASS to construct a Shepard-diagram (see next section).

```{r}
kellogShep <- Shepard(kdist, kellogMDS$points)
plot(kellogShep, pch = 21, cex = 0.8,
     xlab = "d_ij", ylab = "delta and d hat",
     main = "Shepard diagram for the breakfast cereal")
lines(x = kellogShep$x, y = kellogShep$yf, col = "magenta")
```

