---
title: "MDS Assignment 9"
author: "M Meyer (22675760)"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(StatMatch)
library(MASS)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(cluster)
library(combinat)
library(maps)

# setwd("~/Desktop/Multi-Dim Scaling/9")
Africa_MDS <- read.csv("AfricaMDS.csv", header=TRUE, row.names = 1)
Afr_cities <- read.csv("Afr_cities.csv", header=TRUE)
OrdSurvMaps <- read.csv("OrdSurvMaps.csv", header=TRUE, row.names = 1)

speed <- read.csv("Speed.csv", header=TRUE, row.names = 1)
```

## Assignment 9.1

### (a) Implement the algorithm given in C&C section 5.2.2 in an R-function.

```{r}
projProcrustes <- function(X,Y)
{
  Y0 <- Y
  X0 <- X
  # q > p
  q <- ncol(X)
  p <- ncol(Y)
  n <- nrow(X)
  dimdiff <- q-p
  r20 <- 9999999999999999
  r21 <- 99999999
  counter = 0
  
  # 1- padding to ensure the same nr of columns
  while(ncol(Y) < ncol(X))
  {
    Y <- cbind(Y,0)
  }
  
  # continue the process until the new 
  while (r21 < r20)
  {
    counter = counter+1
    r20 <- r21
    # 2- column centre both of them
    Xc <- apply(X, 2, function(x) x - mean(x))
    Yc <- apply(Y, 2, function(y) y - mean(y))
    
    # 3- find svd of Y'X
    YX <- svd(t(Yc)%*%Xc)
      
    L <- diag(YX$d)
    U <- YX$u
    V <- YX$v
    
    # 4- A=VY'
    A <- V%*%t(U)
    
    # Rotate X to XA
    XA <- Xc%*%A
    X <- XA
    # Replace the final q-p columns of Y by final cols of XA
    Y <- cbind(Y0, XA[,-(1:p)])
    
    # Calculate r2
    r2Inner <- t(Y-XA)%*%(Y-XA)
    r2 <- sum(diag(r2Inner))
    r21 <- r2
  }

  
  fittedX <- X[,1:p]
  finalA <- A[,1:p]
  return(list(r2 = r20,fitX = fittedX, A = finalA ,iters = counter))
}

```

\###(b) Use projection Procrustes to project the 3D classical scaling solution in Africa.MDS onto the 2D scatterplot of longitude vs latitude.

```{r}
require(maps)
require(geodist)
capitals <- world.cities[world.cities$capital==1,]
capitals <- capitals[c(3,4,6,28,49,73,80,90,92,94,98,102,109,110,118,124,161,164,223),]
africa2D <- cbind(capitals$lat, capitals$long)

africa.results <- projProcrustes(X = Africa_MDS, Y = africa2D)
mds_df <- as.data.frame(africa.results$fitX)
# Create the plot using ggplot2
sp <- ggplot(mds_df, aes(x = V1, y = V2, label = rownames(mds_df))) +
  geom_point(shape = 16, col = "mediumturquoise") +
  labs(x = "X", y = "Y", title = "Ordinary Projection Procrustes on African Cities")

sp + geom_text( size = 3, hjust=0, vjust=0, col = "turquoise4")
```

### (c) Apply the R-function to the Speed and Ordnance Survey data.

The following function is for the case where p = q.

```{r}
projProcrustes2 <- function(X,Y)
{
  Y0 <- Y
  X0 <- X
  # q > p
  q <- ncol(X)
  p <- ncol(Y)
  n <- nrow(X)
  dimdiff <- q-p
  r20 <- 9999999999999999
  r21 <- 99999999
  counter = 0
  
  # 1- padding to ensure the same nr of columns
  while(ncol(Y) < ncol(X))
  {
    Y <- cbind(Y,0)
  }
  
  # continue the process until the new 
  while (r21 < r20)
  {
    counter = counter+1
    r20 <- r21
    # 2- column centre both of them
    Xc <- apply(X, 2, function(x) x - mean(x))
    Yc <- apply(Y, 2, function(y) y - mean(y))
    
    # 3- find svd of Y'X
    YX <- svd(t(Yc)%*%Xc)
      
    L <- diag(YX$d)
    U <- YX$u
    V <- YX$v
    
    # 4- A=VY'
    A <- V%*%t(U)
    
    # Rotate X to XA
    XA <- Xc%*%A
    X <- XA
    # Replace the final q-p columns of Y by final cols of XA
    # n/a
    
    # Calculate r2
    r2Inner <- t(Y-XA)%*%(Y-XA)
    r2 <- sum(diag(r2Inner))
    r21 <- r2
  }

  fittedX <- X[,1:p]
  finalA <- A[,1:p]
  return(list(r2 = r20,fitX = fittedX, A = finalA ,iters = counter))
}
```

The first plot illustrates the original speed data, the second plot shows how the speed data has been transformed and the third is a plot of the Ordnance Survey map.

```{r}
speed.results <- projProcrustes2(X = as.matrix(speed), Y = as.matrix(OrdSurvMaps))

mds_df3 <- as.data.frame(speed)

sp3 <- ggplot(mds_df3, aes(x = V2, y = V3, label = rownames(mds_df3))) +
  geom_point(shape = 16, col = "coral") +
  labs(x = "X", y = "Y", title = "Original Speed data")
sp3 + geom_text( size = 3, hjust=0, vjust=0, col = "coral4")


mds_df1 <- as.data.frame(speed.results$fitX)

# Create the plot using ggplot2
sp1 <- ggplot(mds_df1, aes(x = V1, y = V2, label = rownames(mds_df1))) +
  geom_point(shape = 16, col = "mediumturquoise") +
  labs(x = "X", y = "Y", title = "Projection Procrustes on Speed data")
sp1 + geom_text( size = 3, hjust=0, vjust=0, col = "turquoise4")


# ORDINANCE data

mds_df2 <- as.data.frame(OrdSurvMaps)

# Create the plot using ggplot2
sp2 <- ggplot(mds_df1, aes(x = V1, y = V2, label = rownames(mds_df2))) +
  geom_point(shape = 16, col = "magenta") +
  labs(x = "X", y = "Y", title = "Ordnance data")
sp2 + geom_text( size = 3, hjust=0, vjust=0, col = "magenta4")

```

## Assignment 9.2

### (a) Find $\mathbf{A}$ and the minimum value of $r^2$ with least squares projection Procrustes.

```{r}
lsProjProcrustes <- function(X,Y)
{
  Y0 <- Y
  X0 <- X
  # q > p
  q <- ncol(X)
  p <- ncol(Y)
  n <- nrow(X)
  dimdiff <- q-p
  YXA0 <- 9999999999
  YXA1 <- 999999999999999
  r2_0 <- 100000000000
  r2_1 <- 0
  counter = 0
  
  # 1- padding to ensure the same nr of columns
  while(ncol(Y) < ncol(X))
  {
    Y <- cbind(Y,0)
  }
  
  # continue the process until the new 
  while ((r2_1-r2_0) > 0.0001)
  {
    counter = counter+1
    # 2- column centre both of them
    Xc <- apply(X, 2, function(x) x - mean(x))
    Yc <- apply(Y, 2, function(y) y - mean(y))
    
    # 3- find svd of Y'X
    YX <- svd(t(Yc)%*%Xc)
      
    L <- diag(YX$d)
    U <- YX$u
    V <- YX$v
    
    # 4- A=VY'
    A <- V%*%t(U)
    
    # 5- Calculate rho
    XAY <- Xc%*%A%*%t(Yc)
    XX <- t(Xc)%*%Xc
    rho <- sum(diag(XAY))/sum(diag(XX))
    
    # Rotate X to XA
    XA <- rho*Xc%*%A
    X <- XA
    # Replace the final q-p columns of Y by final cols of XA
    Y <- cbind(Y0, XA[,-(1:p)])
    
    # Calculate r2
    r2Inner <- t(Y-XA)%*%(Y-XA)
    r2_1 <- sum(diag(r2Inner))
  }


  fittedX <- X[,1:p]
  finalA <- A[,1:p]
  return(list(r2 = r2,fitX = fittedX, A = finalA ,iters = counter))
}
```

```{r}
lsProjProcrustes <- function(X,Y)
{
  Y0 <- Y
  X0 <- X
  # q > p
  q <- ncol(X)
  p <- ncol(Y)
  n <- nrow(X)
  dimdiff <- q-p
  YXA0 <- 9999999999
  YXA1 <- 999999999999999
  counter = 0
  
  # 1- padding to ensure the same nr of columns
  while(ncol(Y) < ncol(X))
  {
    Y <- cbind(Y,0)
  }
  
  # continue the process until the new 
  while (YXA1 > YXA0)
  {
    counter = counter+1
    YXA0 <- YXA1
    # 2- column centre both of them
    Xc <- apply(X, 2, function(x) x - mean(x))
    Yc <- apply(Y, 2, function(y) y - mean(y))
    
    # 3- find svd of Y'X
    YX <- svd(t(Yc)%*%Xc)
      
    L <- diag(YX$d)
    U <- YX$u
    V <- YX$v
    
    # 4- A=VY'
    A <- V%*%t(U)
    
    # Rotate X to XA
    XA <- Xc%*%A
    X <- XA
    # Replace the final q-p columns of Y by final cols of XA
    Y <- cbind(Y0, XA[,-(1:p)])
    
    # Calculate Y'XA
    YXA1 <- sum(diag(t(Y)%*%XA))
  }

  # Calculate r2
  r2Inner <- t(Y-XA)%*%(Y-XA)
  r2 <- sum(diag(r2Inner))
  fittedX <- X[,1:p]
  finalA <- A[,1:p]
  return(list(r2 = r2,fitX = fittedX, A = finalA ,iters = counter))
}

```

```{r}
# X - testee
X <- matrix(c(0.90, 0.83, 0.87, 0.55, 0.56, 0.63, 0.28, 0.38, 0.38,
              -0.09, 0.09, -0.01, 0.79, 0.65, 0.60, 0.27, 0.20, 0.19,
              -0.03, -0.04, 0.07, -0.07, 0.04, 0.03, 0.45, 0.63, 0.77), byrow=FALSE, ncol = 3)
# Y - target
Y <- matrix(c(0.98, 0.76, 0.86, 0, 0, 0, 0, 0, 0,
              0, 0, 0, 1.00, 0.84, 0.77, 0, 0, 0), byrow =FALSE, ncol = 2)

ls.results <- lsProjProcrustes(X=X, Y=Y)


mds_df1 <- as.data.frame(ls.results$fitX)

# Create the plot using ggplot2
sp1 <- ggplot(mds_df1, aes(x = V1, y = V2, label = rownames(mds_df1))) +
  geom_point(shape = 16, col = "cornflowerblue") +
  labs(x = "X", y = "Y", title = "LS Projection Procrustes")
sp1 + geom_text( size = 3, hjust=1.5, vjust=0, col = "blue")
```

### (b) Use the inner project version of projection Procrustes and compare your answer to that obtained in (a).


```{r}
InnerProjProcrustes <- function(X,Y)
{
  Y0 <- Y
  X0 <- X
  # q > p
  q <- ncol(X)
  p <- ncol(Y)
  n <- nrow(X)
  dimdiff <- q-p
  YXA0 <- 9999999999
  YXA1 <- 999999999999999
  counter = 0
  
  # 1- padding to ensure the same nr of columns
  while(ncol(Y) < ncol(X))
  {
    Y <- cbind(Y,0)
  }
  
  # continue the process until the new 
  while (YXA1 > YXA0)
  {
    counter = counter+1
    YXA0 <- YXA1
    # 2- column centre both of them
    Xc <- apply(X, 2, function(x) x - mean(x))
    Yc <- apply(Y, 2, function(y) y - mean(y))
    
    # 3- find svd of Y'X
    YX <- svd(t(Yc)%*%Xc)
      
    L <- diag(YX$d)
    U <- YX$u
    V <- YX$v
    
    # 4- A=VY'
    A <- V%*%t(U)
    
    # Rotate X to XA
    XA <- Xc%*%A
    X <- XA
    # Replace the final q-p columns of Y by final cols of XA
    Y <- cbind(Y0, XA[,-(1:p)])
    
    # Calculate Y'XA
    YXA1 <- sum(diag(t(Y)%*%XA))
  }

  # Calculate r2
  r2Inner <- t(Y-XA)%*%(Y-XA)
  r2 <- sum(diag(r2Inner))
  fittedX <- X[,1:p]
  finalA <- A[,1:p]
  return(list(r2 = r2,fitX = fittedX, A = finalA ,iters = counter))
}
```

```{r}
# X - testee
X <- matrix(c(0.90, 0.83, 0.87, 0.55, 0.56, 0.63, 0.28, 0.38, 0.38,
              -0.09, 0.09, -0.01, 0.79, 0.65, 0.60, 0.27, 0.20, 0.19,
              -0.03, -0.04, 0.07, -0.07, 0.04, 0.03, 0.45, 0.63, 0.77), byrow=FALSE, ncol = 3)
# Y - target
Y <- matrix(c(0.98, 0.76, 0.86, 0, 0, 0, 0, 0, 0,
              0, 0, 0, 1.00, 0.84, 0.77, 0, 0, 0), byrow =FALSE, ncol = 2)

inner.results <- InnerProjProcrustes(X=X, Y=Y)
pp.results <- projProcrustes(X=X, Y=Y)

inner_df <- as.data.frame(inner.results$fitX)
pp_df <- as.data.frame(pp.results$fitX)

ggdat1 <- cbind(rep(1:9,2),rbind(inner_df, pp_df), c(rep("Inner Product PP",9), rep("OPA",9)))
colnames(ggdat1) <- c("id", "V1","V2", "pp.group")

# Create the plot using ggplot2
sp <- ggplot(ggdat1, aes(x = V1, y = V2, label = id)) +
  geom_point(aes(color = factor(pp.group)),shape = 16) +
  labs(x = "X", y = "Y", title = "Inner product & Ordinary Projection Procrustes")

sp + 
  geom_text(aes(color = factor(pp.group)), size = 4, hjust=0, vjust=0)
```
